---
title: "Reproducable Research Assignment1"
author: "IFergus"
date: "March 12, 2015"
output: html_document
---


```{r loadlibraries, echo=TRUE}
library("utils")
library("plyr")
library("xtable")
library("lattice")
```

Download and read the data into a dataframe. 

```{r loaddata, echo=TRUE}

work.file.name <- "activity.csv"
zip.file.name <- "Factivity.zip"

if (!(file.exists(work.file.name))) {
    download.file("https://d396qusza40orc.cloudfront.net/repdata/data/Factivity.zip",
                  destfile = zip.file.name, method = "curl")
    unzip(zip.file.name)
}

original.df <- read.csv(work.file.name)
```


### 1 - Mean total number of steps taken per day 

Using the original dataframe, create a new dataframe (summary.df),  that summarizes the steps based on the day.  Note that observations that have no (NA) steps are excluded.

```{r, echo=TRUE}

summary.df <- ddply(original.df[!(is.na(original.df$steps)), ],
                    .(date),
                    summarise,
                    Mean = round(mean(steps), 2),
                    Median = median(steps),
                    Sum = sum(steps))
```

* Plot a histogram of the total number of steps taken each day

```{r plothistogram, echo=TRUE}

hist(summary.df$Sum,
        main = "Histogram (Steps By Day)",
        xlab = "Steps",
        ylab = "Frequency (Days)")
```


* Display a table showing the mean and median of the total number of steps taken per day

```{r showtable, results='asis'}
print(xtable(summary.df), type = "html")
```


### 2 - The average daily activity pattern


```{r, echo=TRUE}

avg.steps.by.interval.df <- ddply(original.df[!(is.na(original.df$steps)), ],
                                .(interval),
                                summarise,
                                Steps = round(mean(steps), 2))
```

* Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, echo=TRUE}

plot(avg.steps.by.interval.df$interval,
     avg.steps.by.interval.df$Steps,
     xlab="Interval",
     ylab="Steps (Average)",
     type="l")
```

* Determine the 5-minute interval that, on average, contains the maximum number of steps

```{r maxinterval, echo=TRUE}

max.steps.interval <- avg.steps.by.interval.df[avg.steps.by.interval.df$Steps == max(avg.steps.by.interval.df$Steps), ]

```

#### The 5-minute interval that, on average, contains the maximum number of steps is  `r max.steps.interval$interval`.





### 3 - Imputing missing values


* Calculate the mean of all the steps that were provided.  The mean will be used to back-fill those observations for which steps were not provided.  Calulate and store the mean value to mean.of.non.na.steps

```{r calculatemean, echo=TRUE}
mean.of.non.na.steps <- as.integer(mean(original.df$steps[!is.na(original.df$steps)]))
```


* Create a new dataset that is equal to the original dataset but with the missing data filled in. Using the calculated mean value replace hte missing values and create a new dataset called original.backfill.df.


```{r calculate.backfill.mean, echo=TRUE}
original.backfill.df <- original.df
original.backfill.df$steps <- replace(original.backfill.df$steps, is.na(original.backfill.df$steps), mean.of.non.na.steps)
```

* Using the original.backfill.df dataset, create a new dataset (summary.backfill.df),  that summarizes the steps based on the day, after th missing data has been inputted.


```{r calculate.backfill.summary, echo=TRUE}

summary.backfill.df <- ddply(original.backfill.df,
                            .(date),
                            summarise,
                            Mean = round(mean(steps), 2),
                            Median = median(steps),
                            Sum = sum(steps))
```


*  Using the dataset original.backfill.df create a histogram of the total number of steps taken each day -- after missing values were inputted


```{r create.backfill.histogram, echo=TRUE}

hist(summary.backfill.df$Sum,
      main = "Histogram (Steps By Day) -- after missing values were imputed",
      xlab = "Steps",
      ylab = "Frequency (Days)")
```


* Display a table showing the mean and median of the total number of steps taken per day  -- after missing values were inputted 

```{r showtable.with.inputted, results='asis'}
print(xtable(summary.backfill.df), type = "html")
```



### 4 - Are there differences in activity patterns between weekdays and weekends?

* Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r create.type.of.day, echo=TRUE}
original.backfill.df$day.type <- ifelse(weekdays(as.Date(original.backfill.df$date)) %in% c("Saturday","Sunday"), "weekend", "weekday")
original.backfill.df$day.type <- as.factor(original.backfill.df$day.type)
```





```{r create.summaries, echo=TRUE}

summary.weekday.df <- ddply(original.backfill.df[original.backfill.df$day.type == "weekday", ],
                          .(interval),
                          summarise,
                          Steps = round(mean(steps), 2),
                          day = "weekday")

summary.weekend.df <- ddply(original.backfill.df[original.backfill.df$day.type == "weekend", ],
                          .(interval),
                          summarise,
                          Steps = round(mean(steps), 2),
                          day = "weekend")


total.summary.df <- rbind(summary.weekend.df, summary.weekday.df)

```



* Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r create.panel.plots, echo=TRUE}

xyplot(Steps ~ interval | factor(day), total.summary.df, plot.points = FALSE, type = "l", layout = c(1,2))

```
